Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-12 icfp.tex
RM: Edits Sec 2. lets

% \todo[inline]{Cut this one}
In the example from the beginning of
Section~\ref{sec:linearity-semantically} (similar to that towards the end
of Section~\ref{sec:intro}), we discussed a scenario where an
unused let binding can be used to justify a function where its
argument occurs both in the (unused) let-bound expression and in the
body of the let.
Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to
non-strict evaluation often produce
programs with bindings that are selectively used in case alternatives, for
instance:
%
\begin{notyet}
\begin{lstlisting}
f :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f use bool x = let y = use x in case bool of
      True -> x 
      False -> y
\end{lstlisting}
\end{notyet}
%
This example uses \lstinline{x} directly in one
branch and the let-bound \lstinline{y} in the other. Semantically, this program is linear because
the linear resource \lstinline{x} ends up is used exactly once in both case
alternatives, directly in the \lstinline{True} branch or via the
let-bound \lstinline{y} in the \lstinline{False} branch.

Shifting our focus from not using a let binding to using it
more than once.
Despite the binder body only being evaluated
once due to call-by-need, and thus its resources being only used once to compute a result, we can still
only consume the result of the computation once -- perhaps surprisingly, as the
perception so far is that ``resources are consumed during computation'' and
multiple uses of the same let binder share the result that was computed only
once. Illustratively, the following program must \emph{not} typecheck:
%
\begin{noway}
\begin{lstlisting}
f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
f1 use x = let y = use x in (y, y)
\end{lstlisting}
\end{noway}
%
Intuitively, the result of the computation must also be used exactly once,
despite being effectively computed just once, since the result may
still capture (parts of) the linear resource. The trivial example is \lstinline{f1} applied to
\incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
shared. Indeed, if the result of the computation involving the linear resource
was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
resource more than once.
% ROMES:TODO:!!!!
%\todo{this kind of hints that maybe somehow in the
%alternative scrutinizing a "trivial" atom we could make the let binding
%unrestricted after we were sure it was evaluated once, but that isn't easy
%without erasing too many things before, we kind of tried this once}
%
Concretely, the result of evaluating a let binder body using linear resources, if computed, must be
consumed exactly once, or, otherwise, we risk discarding or duplicating said resources.

Lastly, consider a program which defines two let bindings \incode{z} and \incode{y}, where
\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f2 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
\incode{z} isn't evaluated in the let body, and consequently \incode{y} isn't evaluated
either -- never consuming \incode{x}. We use this example to highlight that even for
let bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of \incode{y} as
implying using \incode{x}, and therefore uses of \incode{z} imply using \incode{x}, however, if
neither is used, then \incode{x} isn't used. Since \incode{x} is effectively discarded, this
example also violates linearity.

>>>>

% \todo[inline]{Cut this one}
In the example from the beginning of
Section~\ref{sec:linearity-semantically} (similar to that towards the end
of Section~\ref{sec:intro}), we discussed a scenario where an
unused let binding can be used to justify a function where its
argument occurs both in the (unused) let-bound expression and in the
body of the let.
Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to
non-strict evaluation often produce
programs with bindings that are selectively used in case alternatives, for
instance:
%
\begin{notyet}
\begin{lstlisting}
f :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f use bool x = let y = use x in
  case bool of
    True -> x 
    False -> y
\end{lstlisting}
\end{notyet}
%
This example uses \lstinline{x} directly in one branch and the let-bound
\lstinline{y} in the other. Semantically, as opposed to syntactically, this
program is linear (i.e. linear resources are used \emph{exactly once}) because
the linear resource \lstinline{x} is used exactly once in both case
alternatives, either directly in the \lstinline{True} branch, or \emph{via} the
let-bound \lstinline{y} in the \lstinline{False} branch.

% Shifting our focus from not using a let binding to using it
% more than once.
On the other hand, let us consider a let-binding \lstinline{y} that is used
twice, and that binds an expression that depends on a linear resource
\lstinline{x}.
Despite the binder body only being evaluated once due to call-by-need, and thus
its resources being only used once to compute a result, we can still only
consume the result of the computation once -- perhaps surprisingly, as the
perception so far is that ``resources are consumed during computation'' and
multiple uses of the same let binder share the result that was computed only
once. Illustratively, the following program must \emph{not} typecheck:
%
\begin{noway}
\begin{lstlisting}
f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
f1 use x = let y = use x in (y, y)
\end{lstlisting}
\end{noway}
%
Intuitively, the result of the computation must also be used exactly once,
despite being effectively computed just once, since the result may
still capture (parts of) the linear resource. The trivial example is \lstinline{f1} applied to
\incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
shared. Indeed, if the result of the computation involving the linear resource
was insteaed, e.g., an unrestricted integer, then sharing the result would not
involve consuming the resource more than once.
% ROMES:TODO:!!!!
%\todo{this kind of hints that maybe somehow in the
%alternative scrutinizing a "trivial" atom we could make the let binding
%unrestricted after we were sure it was evaluated once, but that isn't easy
%without erasing too many things before, we kind of tried this once}
%
The result of evaluating a let-binder's body which depends on linear resources,
if computed, must be consumed exactly once, or, otherwise, we risk discarding
or duplicating (the parts of) the resources that were captured in the result.

Lastly, consider a program which defines two let bindings \incode{z} and \incode{y}, where
\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f2 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
\incode{z} isn't evaluated in the let body, and consequently \incode{y} isn't evaluated
either -- never consuming \incode{x}. We use this example to highlight that even for
let-bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of \incode{y} as
implying using \incode{x}, and therefore uses of \incode{z} imply using \incode{x}, however, if
neither is used, then \incode{x} isn't used. Since \incode{x} is effectively discarded, this
example also violates linearity.
